---
title: "**OntoPhylo Tutorial: Character Annotation**"
author: "Diego S. Porto and Sergei Tarasov"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    citation_package: natbib
bibliography: references.bib
biblio-style: "apalike"
---

```{r setup, eval = TRUE, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# STEP 0. Install and load packages.

Before starting this tutorial, please verify if you already have *ontophylo* installed.
If not, then you can install it by running the code below:
```{r eval = FALSE, include = TRUE, message = FALSE, warning = FALSE}
remotes::install_github("diegosasso/ontophylo")
```

Now, let's load the package.
```{r eval = TRUE, include = TRUE, message = FALSE, warning = FALSE}
library(ontophylo)
```

Let's also load some other packages that will be useful in this tutorial. If you have not them installed, please do so by running `install.packages()`.
```{r eval = TRUE, include = TRUE, message = FALSE, warning = FALSE}
library(ontoFAST)
library(ontologyIndex)
library(tidyverse)
```

# STEP 1. Import data.

For this tutorial and the following ones, we will use a modified data set from \citet{sharkey2012}.
The modified data set contains 239 morphological characters for 87 species of Hymenoptera.
```{r eval = TRUE, include = TRUE, message = FALSE, warning = FALSE}
# Import character matrix.
hym_mat <- readRDS("data/hym_matrix.RDS")
```

# Overview on character annotation.

Each phylogenetic character is a statement about a particular phenotype observed in one or more organisms.
In its simplest form, a character statement is an affirmation about the absence/presence or quality of an anatomical entity. More complex statements may relate two or more anatomical entities (e.g. entity A connected to entity B) or include relative measurements (e.g. entity A wider than long).

Character statements are usually represented as strings of text in natural language (=written for humans).
In such format, they cannot be easily parsed or reasoned by computers. A simple alternative to make them computer-parsable is to convert character statements to semantic statements by using a structured data model and standardized syntax. 

The Entity-Quality (EQ) syntax represents character statements as pairs of ontology terms, one from an anatomy ontology and another from a quality ontology. For example, the character statement "Mandible shape: curved" describing the phenotype of a hymenopteran insect can be represented as `Mandible` ([HAO_0000506](http://purl.obolibrary.org/obo/HAO_0000506)): `curved` ([PATO_0000406](http://purl.obolibrary.org/obo/PATO_0000406)) by linking an anatomy term from the Hymenoptera Anatomy Ontology (HAO) \citep{yoder2010} to a quality term from the Phenotype and Trait Ontology (PATO) \citep{gkoutos2005}.

For the purposes of the PARAMO pipeline \citep{tarasoval2019}, only the anatomy term is necessary to guide character amalgamation.
We just need to link each character statement (i.e. a column in a character matrix) to a specific term from an anatomy ontology that better defines the morphological structure under consideration. This procedure is called 'annotation'. Annotation can be facilitated by using semi-automatic methods such as those implemented in *ontoFAST* \citep{tarasoval2022} but usually still requires expert-based input to make the final decision. In this tutorial, we will show how to employ pre-annotated characters or semi-automatic annotations to query the ontology.

# STEP 2A. Querying the ontology with pre-annotated characters.

Let's start with pre-annotated characters. First, import the csv table with pre-cooked ontology annotations running the code below: 
```{r eval = TRUE, include = TRUE, message = FALSE, warning = FALSE}
# Import annotations.
hym_annot <- read_csv(file = "data/hym_annot.csv")
```

To guide the character amalgamations in the PARAMO pipeline, we need first to cluster characters based on their ontology annotations. In this tutorial, we will use the Hymenoptera Anatomy Ontology. Terms from this ontology were used to manually annotate all character statements, as seen in the `hym_annot` object. In the following chunk of code, we will first import the HAO ontology in OBO format using the package *ontologyIndex* \citep{greene2017}. Then we will make a query to retrieve all characters annotated with terms that are *part_of* three main anatomical regions of the Hymenoptera anatomy: head, mesosoma and metasoma.
```{r eval = TRUE, include = TRUE, message = FALSE, warning = FALSE}
# Import HAO.
onto <- get_OBO("data/HAO.obo", extract_tags = "everything", 
                propagate_relationships = c("BFO:0000050", "is_a"))

# Define query terms for anatomical regions.
query_anat <- c("head", "mesosoma", "metasoma")

# Set a table of matches between characters and ontology terms.
char_info <- hym_annot %>% select(char_id, onto_id)

# Group characters by anatomical regions.
HYM_ANAT <- RAC_query(char_info, onto, query_anat)
```

# STEP 2B. Querying the ontology using *ontoFAST*.

Note that in the previous step, annotations to ontology terms were given in the `onto_id` column. 
```{r eval = TRUE, include = TRUE, message = FALSE, warning = FALSE}
hym_annot %>% slice(1:8)
```

However, most real-world phylogenetic data sets are not annotated with ontology terms. The only information often available are the character statements written in natural language, for example, as annotations in the character block of a NEXUS file. In our example data set, character statements (not including state descriptions) are shown in the `char` column. Let's try to match some of these with terms from HAO using the *ontoFAST* semi-automatic approach.
```{r eval = TRUE, include = TRUE, message = FALSE, warning = FALSE}
# Pre-organize the ontology object.
onto$parsed_synonyms <- syn_extract(onto)
onto$id_characters <- hym_annot$char_id
onto$name_characters <- setNames(hym_annot$char,hym_annot$char_id)

# Run ontoFAST.
auto_annot <- annot_all_chars(onto)
```

Let's check the candidate terms recovered with *ontoFAST* for a small sample of character statements.
```{r eval = TRUE, include = TRUE, message = FALSE, warning = FALSE}
cand_annot <- setNames(auto_annot[6:8], hym_annot$char[6:8])
lapply(cand_annot, function(x) onto$name[names(onto$name) %in% x] )
```

Although *ontoFAST* does a great job in finding good matches, there is still the need for an expert in the group to select the terms that best apply to the anatomical entity described in each character statement. Once this is done (we are not going to annotate +200 characters!), a table similar to `hym_annot` can be constructed and used to query the ontology, as demonstrated before.


Finally, we will sample 10 characters from each anatomical region for further analyses in the following tutorials and save the results.
```{r eval = TRUE, include = TRUE, message = FALSE, warning = FALSE}
# Get a sample of characters from each anatomical region.
set.seed(42)
HYM_ANAT <- lapply(HYM_ANAT, function(x) sample(x, 10))
HYM_ANAT

# Create a folder to store RData files.
dir.create("RData")

# Save workspace.
save.image("RData/step1_annot.RData")
```
